Fractional Types

### 2016-12-19 typespp.tex
Minor copy-editing

If we are to build a compositional programming language around iteration and
symmetry groupoids, we need to ensure that they compose sensibly with the
existing type formers. Groupoids, viewed as categories, come associated with
natural notions of sums and products, and one might expect or hope that
identities which hold for rational numbers lift to identities in our
situation. The most prominent and important identity we seek is:
\[
\order{p} \circledast \iorder{p} \simeq \order{\idiso}
\]
The left hand side is the categorical product of the iteration groupoid and
symmetry groupoid for some arbitrary 1-combinator $p$. The cardinality of this
groupoid is 1. The right hand side is the iteration groupoid for the identity
1-combinator which also has cardinality 1. The groupoids on either side are
however not categorically equivalent and hence $\simeq$ cannot be categorical
equivalence.  There are however other notions of groupoid equivalence, such as
Morita equivalence~\cite[C5.3]{johnstone2002sketches}~\cite{looping}, that
identify the two groupoids above.  Although the notion of equivalence we use in
the next section appears novel, it is related ``in spirit'' to Morita
equivalence and we find it useful to give a small example. We will verify that
the product of the groupoid $\order{\permfive}$ of cardinality 3 and the
groupoid $\iorder{\permfive}$ of cardinality $\frac{1}{3}$ is equivalent to the
trivial groupoid with only one distinct object and one trivial automorphism. To
avoid clutter we assume the groupoids are strict, e.g., that the groupoid
$\order{\permfive}$ has three elements (instead of 3 isomorphism classes of
elements). The definition of \emph{Morita equivalence} in this case reduces to
to finding a \emph{Morita morphism} from $\order{\idiso}$ to
$\order{\permfive} \circledast \iorder{\permfive}$. This is a functor $\psi$
from the unit groupoid to the product groupoid, say $G$, with an additional
condition that the commutative diagram in figure~\ref{fig:morita} is a
\emph{pullback}, where $s,t$ are the \emph{source} and \emph{target} maps, and
$\ot$, $\mathbb{3}$ are the $1$- and $3$-point discrete groupoids defined
previously.  The intuition is that the codomain consists of three objects with
associated morphisms, each of them counting as a ``third.'' The functor from
$\order{\idiso}$ to $\order{\permfive} \circledast \iorder{\permfive}$ must
include an arbitrary choice of ``which third'' to target and the universality
condition of the pullback ensures that the choice is ultimately insignificant as
it leads to the same result as every other possible functor. To summarize the
equivalence between the two groupoids requires a \emph{local} map that chooses a
target and a \emph{global} condition ensuring that the choice ultimately
annihilates. The challenge we are facing is to turn this idea into an
operational semantics that somehow combines the local and global aspects of the
equivalence. We will be inspired by the folklore result that ``substitution into
a dependent type is a pullback''~\cite{subpullback} (but which turns out to have
been fully explained in~\cite{practicalfoundations}) as explained in the next
section.

>>>>

If we are to build a compositional programming language around iteration and
symmetry groupoids, we need to ensure that they compose sensibly with the
existing type formers. Groupoids, viewed as categories, come associated with
natural notions of sums and products, and one might expect or hope that
identities which hold for rational numbers lift to identities in our
situation. The most prominent and important identity we seek is:
\[
\order{p} \circledast \iorder{p} \simeq \order{\idiso}
\]
The left hand side is the categorical product of the iteration groupoid and
symmetry groupoid for some arbitrary 1-combinator $p$. The cardinality of this
groupoid is 1. The right hand side is the iteration groupoid for the identity
1-combinator which also has cardinality 1. The groupoids on either side are
however not isomorphic, there do not exist full and faithful functors between
them, and hence $\simeq$ cannot be categorical equivalence.

There are however weaker notions of groupoid equivalence, such as \emph{Morita
  equivalence}~\cite[C5.3]{johnstone2002sketches}~\cite{looping}, that identify
the two groupoids above.  Although the notion of equivalence we use in the next
section appears novel, it is related ``in spirit'' to Morita equivalence and we
find it useful to give a small example. We will verify that the product of the
groupoid $\order{\permfive}$ of cardinality 3 and the groupoid
$\iorder{\permfive}$ of cardinality $\frac{1}{3}$ is Morita equivalent to the
trivial groupoid with only one distinct object and one trivial automorphism. To
avoid clutter, we assume the groupoids are strict, e.g., that the groupoid
$\order{\permfive}$ has three elements (instead of 3 isomorphism classes of
elements). Getting an equivalence in this case reduces to finding a \emph{Morita
  morphism} from $\order{\idiso}$ to
$\order{\permfive} \circledast \iorder{\permfive}$. This is a functor $\psi$
from the unit groupoid to the product groupoid, say $G$, with an additional
condition that the commutative diagram in figure~\ref{fig:morita} is a
\emph{pullback}, where $s,t$ are the \emph{source} and \emph{target} maps, and
$\ot$, $\mathbb{3}$ are the $1$- and $3$-point discrete groupoids defined
previously.  The intuition is that the codomain consists of three objects with
associated morphisms, each of them counting as a ``third.'' The functor from
$\order{\idiso}$ to $\order{\permfive} \circledast \iorder{\permfive}$ must
include an arbitrary choice of ``which third'' to target and the universality
condition of the pullback ensures that the choice is ultimately insignificant as
it leads to the same result as every other possible functor.

To summarize, the equivalence between the two groupoids requires a \emph{local}
map that chooses a target and a \emph{global} condition ensuring that the choice
ultimately annihilates. The challenge we are facing is to turn this idea into an
operational semantics that somehow combines the local and global aspects of the
equivalence. We will be inspired by the folklore result in categorical semantics
of dependent type theory, that ``substitution into a dependent type is a
pullback''~\cite{subpullback}, (but which turns out to have been fully explained
in~\cite{practicalfoundations}) as explained in the next section.
